Nodal Discontinuous Galerkin Method for the Two-Moment Model of Neutrino Transport

### 2017-03-21 Documents/M1/dgM1Curvilinear.tex
Updates to documents

To connect the elements in Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\bcF}^{i}(\bcM_{\DG})$ is a numerical flux approximating the flux on the $i$th surface of $\bK$.  
For this purpose we define the numerical flux function $\vect{f}^{i}$, which evaluates the numerical flux given values from both sides of an element interface; i.e.,
\begin{equation}
  \widehat{\bcF}^{i}(\bcM_{\DG})=\vect{f}^{i}(\bcM_{\DG}(x^{i,-},\tilde{\bx}^{i}),\bcM_{\DG}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$, e.g., in the arguments of $\bcM_{\DG}(x^{i,-/+},\tilde{\bx}^{i})$, indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
For example, the simple Lax-Friedrichs flux, which we use in this paper, is given by
\begin{equation}
  \vect{f}^{i}(\bcM^{-},\bcM^{+})
  =\f{1}{2}\,\big(\,\bcF^{i}(\bcM^{-})+\bcF^{i}(\bcM^{+})-\alpha^{i}\,(\,\bcM^{+}-\bcM^{-}\,)\,\big),
\end{equation}
where $\alpha^{i}=||\mbox{eig}\big(\partial\bcF^{i}/\partial\bcM\big)||_{\infty}$ is the largest eigenvalue of the flux jacobian.  
For classical neutrinos, which propagate at the speed of light, we take $\alpha^{i}=1$.

@@

\section{Semi-Implicit Time Integration}

For time integration we employ the method lines.  
Having specified the spatial discretization for the advection terms, we consider the discretized moment equations in Eq.~\eqref{eq:semidiscreteDiscretized} as a system of ordinary differential equations (ODEs) to evolve forward in time, and write
\begin{equation}
  d_{t}\bcM_{\bk}
  =\bcF_{\bx}(\bcM)_{\bk}+\bcC(\bcM)_{\bk},
  \label{eq:momentEquationsODEs}
\end{equation}
where the operator $\bcF_{\bx}:\bbR^{d+1}\to\bbR^{d+1}$ represents the discretization of the spatial divergence operator.

@@

We benchmark the DG implementation against a series of test problems.  
We will vary the degree $k$ of the polynomial representation used in DG method from 0 to 3, which we then denote DG($k$).  
First we consider test problems in the streaming limit.  
In this case we use the explicit SSP-RK time integrators.  
The 2-stage and 3-stage methods are denoted SSP-RK2 and SSP-RK3, respectively.  
As an example, an expected third-order accurate method (for problems with smooth solutions), using DG(2) and SSP-RK3, will then be denoted DG(2)+RK3.

@@

This problem is taken from \citet{pons_etal_2000}

@@

\begin{figure}
  \begin{center}
    \includegraphics[scale=0.5]{./Figures/RiemannProblem1D}
  \end{center}
  \caption{Results from the one-dimensional Riemann problem in Eq.~\eqref{eq:riemannProblem}, obtained with the DG(2)+RK3 method using $240$ elements, the HLL Riemann solver, and the Eddington factor in Eq.~\eqref{eq:eddingtonFactorLevermore}.  
  The number density $\cJ$ (black) and the number flux density $\cH_{r}$ (red) are plotted for $t=0.1$.}
  \label{fig:riemannProblem}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \includegraphics[scale=0.32]{./Figures/HomogeneousSphere1D_Chi_1e-4_Density} &
      \includegraphics[scale=0.32]{./Figures/HomogeneousSphere1D_Chi_1e-6_Density} \\
      \includegraphics[scale=0.32]{./Figures/HomogeneousSphere1D_Chi_1e-4_FluxFactor} &
      \includegraphics[scale=0.32]{./Figures/HomogeneousSphere1D_Chi_1e-6_FluxFactor} \\
      \includegraphics[scale=0.32]{./Figures/HomogeneousSphere1D_Chi_1e-4_EddingtonFactor} &
      \includegraphics[scale=0.32]{./Figures/HomogeneousSphere1D_Chi_1e-6_EddingtonFactor}
    \end{tabular}
  \end{center}
  \caption{Results from solving the Homogeneous sphere problem with the DG(2)+SIRK2 method using $100$ elements.  
  In the left panels, the absorption opacity is set to $\chi_{0}=10^{-4}$~cm$^{-1}$, while in the right panels it is set to $\chi_{0}=10^{-6}$~cm$^{-1}$.  
  The numerical results (dashed lines) are compared to the analytical solution (solid lines).}
  \label{fig:homogeneousSphere1D}
\end{figure}

@@

>>>>

To connect the elements in Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\bcF}^{i}(\bcM_{\DG})$ is a numerical flux approximating the flux on the $i$th surface of $\bK$.  
For this purpose we define the numerical flux function $\vect{f}^{i}$, which evaluates the numerical flux given values from both sides of an element interface; i.e.,
\begin{equation}
  \widehat{\bcF}^{i}(\bcM_{\DG})=\vect{f}^{i}(\bcM_{\DG}(x^{i,-},\tilde{\bx}^{i}),\bcM_{\DG}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$, e.g., in the arguments of $\bcM_{\DG}(x^{i,-/+},\tilde{\bx}^{i})$, indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
For example, the simple Lax-Friedrichs (LF) flux, which we have implemented, is given by
\begin{equation}
  \vect{f}^{\mbox{\tiny LF},i}(\bcM^{-},\bcM^{+})
  =\f{1}{2}\,\big(\,\bcF^{i}(\bcM^{-})+\bcF^{i}(\bcM^{+})-\alpha^{i}\,(\,\bcM^{+}-\bcM^{-}\,)\,\big),
\end{equation}
where $\alpha^{i}=||\mbox{eig}\big(\partial\bcF^{i}/\partial\bcM\big)||_{\infty}$ is the largest eigenvalue of the flux jacobian.  
For classical neutrinos, which propagate at the speed of light, we take $\alpha^{i}=1$ (i.e., the global LF flux).  
We have also implemented the Harten-Lax-van Leer (HLL) flux
\begin{equation}
  \vect{f}^{\mbox{\tiny HLL},i}(\bcM^{-},\bcM^{+})
  =
\end{equation}

@@

\section{Time Integration}

For time integration we employ the method lines.  
Having specified the spatial discretization for the advection terms and the representation of the collision terms, we consider the discretized moment equations in Eq.~\eqref{eq:semidiscreteDiscretized} as a system of ordinary differential equations (ODEs) to evolve forward in time, and write
\begin{equation}
  d_{t}\bcM_{\bk}
  =\bcF_{\bx}(\bcM)_{\bk}+\bcC(\bcM)_{\bk}
  \equiv\bcL(\bcM)_{\bk},
  \label{eq:momentEquationsODEs}
\end{equation}
where the operator $\bcF_{\bx},\bcC:\bbR^{d+1}\to\bbR^{d+1}$ represents the discretization of the spatial divergence operator and the collision operator modeling neutrino-matter interactions, respectively.  
(In the following, to simplify notation, we drop the spatial subscript $\bk$, but note that a system of ODEs is to solved for each spatial point.)

\subsection{Explicit Integration Scheme: SSP-RK Methods}
\label{sec:explicitTime}

For non-stiff problems, we employ the strong stability-preserving Runge-Kutta (SSP-RK) methods \citep[e.g.,][]{gottlieb_etal_2001}, which, for evolving the moments from $t^{n}$ to $t^{n+1}=t^{n}+\dt$ ($\bcM^{n}\to\bcM^{n+1}$), take the general form
\begin{align}
  \bcM^{(0)}
  &=\bcM^{n}, \\
  \bcM^{(i)}
  &=\sum_{k=1}^{i-1}\alpha_{ik}\,\big[\,\bcM^{(k)}+\beta_{ik}\,\dt\,\bcL(\bcM^{(k)})\,\big], \quad i=1,\ldots,m 
  \label{eq:ssprkStages} \\
  \bcM^{n+1}
  &=\bcM^{(m)}.  
\end{align}
For second- and third-order temporal accuracy (denoted SSP-RK2 and SSP-RK3, respectively), the coefficient matrices $\alpha_{ik}$ and $\beta_{ik}$ are given in Table~\ref{tab:ssprkCoefficients}.  
Note that $\alpha_{ik},\beta_{ik}\ge0$ and $\sum_{k}\alpha_{ik}=1~\forall i$, so that the RK stages in Eq.~\eqref{eq:ssprkStages} are simply convex combinations of forward Euler steps with time step $\beta_{ik}\,\dt$ \citep{shuOsher_1988}.

\begin{table}
  \begin{center}
  \caption{Optimal coefficients for SSP-RK Methods \citep{shuOsher_1988} \label{tab:ssprkCoefficients}}
  \begin{tabular}{cccccccccccccccccc}
    \midrule
     SSP-RK2 & $\alpha_{ik}$ & & & & $\beta_{ik}$ & & & SSP-RK3 & $\alpha_{ik}$ & & & & $\beta_{ik}$ \\
    \midrule
    \midrule
      & $1$ & & & & 1 & & & & 1 & & & & 1 \\  
      & $1/2$ & $1/2$ & & & $0$ & $1$ & & & $3/4$ & $1/4$ & & & 0 & 1  \\
      & & & & & & & & & $1/3$ & 0 & $2/3$ & & 0 & 0 & 1 \\ 
    \midrule
    \midrule
  \end{tabular}
  \end{center}
\end{table}

\subsection{Semi-Implicit Runge-Kutta Methods}
\label{sec:semiImplicitTime}

For stiff problems, the explicit methods in Section~\ref{sec:explicitTime} impose a severe restriction on the time step for numerical stability.  
In essence, the time step must be chosen small enough to resolve the collision time (the time between subsequent neutrino-matter interactions), which is much shorter than the streaming time (induced by $\bcF_{\bx}$) in regions of dense nuclear matter in a proto-neutron star.

To overcome this time step restriction, we integrate the collision term with implicit methods.  
Following \citet{chertock_etal_2015}, we define the forward Euler stages in Eq.~\eqref{eq:ssprkStages}
\begin{equation}
  \bcM_{ik}^{\mbox{\tiny FE}}
  =\bcM^{(k)}+\beta_{ik}\,\dt\,\big(\,\bcF_{\bx}(\bcM^{(k)})+\bcC(\bcM^{(k)})\,\big).  
\end{equation}
The semi-implicit Runge-Kutta (SIRK) method is obtained by substituting $\bcM_{ik}^{\mbox{\tiny FE}}\to\bcM_{ik}^{\mbox{\tiny SI}}$, where $\bcM_{ik}^{\mbox{\tiny SI}}$ is obtained by solving the implicit equation
\begin{equation}
  \bcM_{ik}^{\mbox{\tiny SI}}
  =\bcM^{(k)}+\beta_{ik}\,\dt\,\big(\,\bcF_{\bx}(\bcM^{(k)})+\bcC(\bcM_{ik}^{\mbox{\tiny SI}})\,\big).  
\end{equation}
As discussed by \citet{chertock_etal_2015}, the time-integration method using Eq.~\eqref{eq:ssprkStages}, with $\bcM_{ik}^{\mbox{\tiny FE}}$ replaced by $\bcM_{ik}^{\mbox{\tiny SI}}$, is at most first-order accurate in time.  
To increase the temporal accuracy to second order, \citet{chertock_etal_2015} propose a correction step to be applied after the last stage.  
We defer the development of the correction step, and use the first-order accurate time integrator for stiff problems.  
We note that the SIRK method reduces to the SSP-RK methods in Section~\ref{sec:explicitTime} when $\bcC=0$.

@@

We benchmark the DG implementation against a series of test problems.  
We will vary the degree $k$ of the polynomial representation used in DG method from 0 to 3, which we then denote DG($k$).  
First we consider test problems in the streaming limit.  
In this case we use the explicit SSP-RK time integrators in Section~\ref{sec:explicitTime}.  
The 2-stage and 3-stage methods are denoted SSP-RK2 and SSP-RK3, respectively.  
As an example, an expected third-order accurate method (for problems with smooth solutions), using DG(2) and SSP-RK3, will then be denoted DG(2)+RK3.

@@

This problem in spherical symmetry is taken from \citet{pons_etal_2000}, and consists of a gaussian-shaped wave propagating in the radial direction.  
The computational domain is given by $D=\{r:r\in[0.2,10.2]\}$.  
The analytical solution is given by
\begin{equation}
  \cJ(r,t)=\cH_{r}(r,t)
  =\exp\big[-\big(r-t\big)^{2}\big]/r^{2}.  
\end{equation}
We let the boundary conditions be given by the analytical solution, and integrate in time from $t=0$ to $t=7$.

Snapshots of the solution ($\cJ$ versus $r$), obtained with the DG(2)+RK3 scheme, are shown in Figure~\ref{fig:sphericalWave} for $t=\{2.5,3.5,5.0,7.0\}$ (open circles).  
For comparison, the analytical solution is plotted with dotted lines.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \includegraphics[scale=0.4]{./Figures/RiemannProblem1D} &
      \includegraphics[scale=0.4]{./Figures/RiemannProblem1D_Limiters}
    \end{tabular}
  \end{center}
  \caption{Results from the one-dimensional Riemann problem in Eq.~\eqref{eq:riemannProblem}, obtained with the DG(2)+RK3 method using $240$ elements, the HLL Riemann solver, and the Eddington factor in Eq.~\eqref{eq:eddingtonFactorLevermore}.  
  The number density $\cJ$ (black) and the number flux density $\cH_{r}$ (red) are plotted for $t=0.1$.}
  \label{fig:riemannProblem}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \includegraphics[scale=0.4]{./Figures/HomogeneousSphere1D_Chi_1e-4_Density} &
      \includegraphics[scale=0.4]{./Figures/HomogeneousSphere1D_Chi_1e-6_Density} \\
      \includegraphics[scale=0.4]{./Figures/HomogeneousSphere1D_Chi_1e-4_FluxFactor} &
      \includegraphics[scale=0.4]{./Figures/HomogeneousSphere1D_Chi_1e-6_FluxFactor} \\
      \includegraphics[scale=0.4]{./Figures/HomogeneousSphere1D_Chi_1e-4_EddingtonFactor} &
      \includegraphics[scale=0.4]{./Figures/HomogeneousSphere1D_Chi_1e-6_EddingtonFactor}
    \end{tabular}
  \end{center}
  \caption{Results from solving the Homogeneous sphere problem with the DG(2)+SIRK2 method using $100$ elements.  
  In the left panels, the absorption opacity is set to $\chi_{0}=10^{-4}$~cm$^{-1}$, while in the right panels it is set to $\chi_{0}=10^{-6}$~cm$^{-1}$.  
  The numerical results (dashed lines) are compared to the analytical solution (solid lines).}
  \label{fig:homogeneousSphere1D}
\end{figure}

@@

\begin{table}
  \caption{Parameters for homogeneous sphere problem with tabulated opacities.}
  \begin{tabular}{ccccc}
    Model & $\rho$ [g~cm$^{-3}$] & $T$ [MeV] & $Y_{e}$ & $\mu_{\nu}$ [MeV] \\
    \midrule \midrule
    A & $1.0\times10^{14}$ & 21.0  & 0.25 &   90.65 \\
    B & $1.0\times10^{13}$ & 16.0 & 0.14 &     4.80 \\
    C & $1.0\times10^{12}$ &   8.0 & 0.12 & -   0.57 \\
    D & $1.0\times10^{11}$ &   8.0 & 0.15 & - 10.87 \\
    \midrule \midrule
  \end{tabular}
\end{table}

\subsection{Diffuse Scattering}

\subsection{Deleptonization}
